The Lipid Interactome

### 2025-02-05 Background/ProteomicsUsingMultifunctionalProbes.qmd
EOD, updates to overview pages

---
title: "Background: Using multifunctional probes for proteomics"
---

@@

::: border
![](/Background/BaselineProteomics_ExperimentalDesign.png){group="Overview_gallery"}
:::

@@

@@

@@

@@

#### Quantitative Mass Spectrometry with Tandem Mass Tagging

Tandem Mass Tag (TMT) labeling is a powerful quantitative mass spectrometry technique that enables the simultaneous analysis of multiple samples in a single MS run. This approach improves efficiency and allows for precise relative quantification of proteins across experimental conditions.

>>>>

---
title: "Background: Using multifunctional probes for proteomics"
format: 
  html:
    embed-resources: false
    toc: true
    toc-depth: 4
  docx:
    fig-height: 5
    reference-doc: /Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/Resources/custom-reference-doc.docx
---

@@

In this overview, we aim to explain the fundamental procedures of utilizing lipid probes in proteomics analysis. [@fig-ExperimentalSchema] depicts the steps described below

::: border
![](/Background/BackgroundFigures/BaselineProteomics_ExperimentalDesign.png){#fig-ExperimentalSchema group="Overview_gallery"}
:::

@@

As has been reported previously by Höglinger et al. (2017), Muller et al. (2021), Farley et al. (2024), Farley et al. (2024), and Thomas et al. (2025), lipid probes are readily taken up by cell membranes within minutes upon treatment with lipid probes at concentrations ranging from 2 \muM (as for trifunctional sphingosine and sphinganine) to 5 \muM (as for trifunctional phosphatidic acid and phosphatidylethanolamine) to 100 \muM (as for trifunctional diacylglycerol)[@hoglingerTrifunctionalLipidProbes2017; @mullerSynthesisCellularLabeling2021; @farleyTrifunctionalFattyAcid2024; @farleyTrifunctionalSphinganineNew2024a; @thomasTrifunctionalLipidDerivatives2025]. These studies utilized the inherent fluorescence of the coumarin cage to validate the uptake and wide intercalation of the lipid probes throughout the cell[@schultzChemicalToolsLipid2023].

### Uncaging

- Irradiation with 405 nm wavelength light induces the release of the coumarin photocage and thus allows for the recognition and metabolism of the headgroup of the lipid probe.

- Substantial work has validated that these lipid probes are recognized by the cell as their native counterparts, as the remaining functional groups have minimal effect on the probes'  biophysical properties. This has been validated in numerous studies, including Höglinger et al. (2015) and Höglinger et al. (2017), Muller et al. (2020) and Muller et al. (2020), and Farley et al. (2024) and Farley et al. (2024)[@hoglingerIntracellularSphingosineReleases2015a; @hoglingerTrifunctionalLipidProbes2017; @mullerSynthesisCellularLabeling2020; @mullerSynthesisCellularLabeling2021].

### Crosslinking

- Irradiation with 355 nm wavelength light ()

@@

- Cell lysis

- Affinity purification of probe-conjugated proteins
  * Example using azide beads: Perform click chemistry reaction in presence of beads coated with picolyl-azide functional groups for covalent capture
  * Conjugates of protein-lipid-bead are covalently bound and can thus be stringently washed with detergent solutions to remove non-specific proteins. Detergent can then be removed with repeated washes of 8 M urea and 20 % acetonitrile. 
* Captured proteins can be reduced and alkylated on-bead. 
* Peptides can be digested off the protein-lipid-bead conjugates via tryptic digest. Digested peptides are then desalted to remove any remaining contaminants using reverse phase desalting columns.

@@

Broadly, all mass spectrometry techniques for proteomic identification of differentially-enriched proteins will involve injection into a mass spectrometer via liquid chromatography. Within the mass spectrometer, individual peptides can be isolated and fragmented to determine their sequence.

There are several methodologies commonly used to quantify mass spectrometry data. At the highest level, these may be separated into label-free and label-based quantification techniques. Both categories are valid for reporting differential enrichment, and each has pros and cons -- discussed in detail below.

@@

#### Label-based quantification

All label-based mass spectrometry quantification methods use stable isotopes to distinguish the source of proteins within a pooled collective of several experimental samples. Previously, the technique known as Stable Isotope Labeling with Amino acids in Cell Culture (SILAC) was common, though this has fallen out of use due to cost and difficult sample preparation.

Modern label-based techniques such as iTRAQ (isotopic Tagging for Relative and Absolute  Quantification) and TMT (Tandem Mass Tagging) have become the norm. For the purposes of this overview, we will focus on TMT as it has greater capacity for samples per experiment (16+ as opposed to 8), and the techniques are fundamentally similar.

TMT labeling is a powerful quantitative mass spectrometry technique that enables the simultaneous analysis of multiple samples in a single MS run. This approach improves efficiency and allows for precise relative quantification of proteins across experimental conditions.

### 2025-02-05 BioinformaticsApplicationNote/GG_2025_applicationNote.qmd
EOD, updates to overview pages

Recent advances in developing multifunctionalized lipid analogs have revolutionized the field, enabling the systematic interrogation of lipid interactomes. These probes incorporate strategically designed functional groups that enable covalent crosslinking, affinity enrichment, and controlled activation. Most commonly, these include: (1) a diazirine ring for photoactivatable crosslinking upon exposure to 355 nm UV light, stabilizing lipid–protein interactions; (2) a terminal alkyne, permitting click chemistry-based enrichment; and, in some cases, (3) a coumarin photocage, which shields the lipid probe from premature enzymatic modification until selectively released by 405 nm light. Together, these modifications enable direct proteomic identification of lipid interactors, distinguishing true binding partners from transient or non-specific associations[@haberkantFatFabulousBifunctional2014; @schultzChemicalToolsLipid2023; @farleyTrifunctionalSphinganineNew2024a].

@@

-   Column descriptions:
    -   Gene name or Accession Number
    -   Enrichment value (Experimental versus Control fold-change, typically log2-transformed)
    -   p value: the results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis testing
    -   AveExpr: An averaged value of the m/z ion intensity of the Experimental condition and Control condition used for the enrichment value (used to prepare MA plots for quality control)
    -   False-discovery rate of protein identification
-   We are working on providing visualization functionality for studies which utilized Peptide Spectral Matching to report enrichment toward lipid probes.

@@

- What we hope this resource enables researchers to do:
  - Explore previously-published results and guide ongoing research.
  - Orthogonally verify and characterize the interactions of bioactive lipid species and enriched proteins.
-   Longevity – this site is designed to expand as more researchers utilize functionalized lipid probes as targets for interactomics.
  - We plan to incorporate a section for each lipid interactome in which validating studies may be referenced -- allowing rapid cross-referencing of research which has already been done.
  - Data submission - with the provided [data submission format](/ContactUs/DataSubmission.qmd), researchers can submit their data for inclusion in this database.

## Abbreviations
MS

>>>>

Recent advances in developing multifunctionalized lipid analogs have revolutionized the field, enabling the systematic interrogation of lipid interactomes. These probes incorporate strategically designed functional groups that enable covalent crosslinking, affinity enrichment, and controlled activation. Most commonly, these include: (1) a diazirine ring for photoactivatable crosslinking upon exposure to 355 nm ultraviolet (UV) light, stabilizing lipid–protein interactions; (2) a terminal alkyne, permitting click chemistry-based enrichment; and, in some cases, (3) a coumarin photocage, which shields the lipid probe from premature enzymatic modification until selectively released by 405 nm light. Together, these modifications enable direct proteomic identification of lipid interactors, distinguishing true binding partners from transient or non-specific associations[@haberkantFatFabulousBifunctional2014; @schultzChemicalToolsLipid2023; @farleyTrifunctionalSphinganineNew2024a].

@@

-   Column descriptions:
    -   *Gene name or Accession Number*
    -   *Enrichment value*: Fold change of Experimental versus Control fold-change, typically log2-transformed -- ideally the product of TMT or other quantitative mass spectrometry methods.
    -   *p value*: the results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis testing
    -   *AveExpr*: An averaged value of the m/z ion intensity of the Experimental condition and Control condition used for the enrichment value (used to prepare MA plots for quality control)
    -   *FDR*: False-discovery rate of protein identification
-   We are working on providing visualization functionality for studies which utilized Peptide Spectral Matching to report enrichment toward lipid probes.

@@

The advent of functionalized lipid probes has fundamentally transformed lipid biology by enabling the proteomic identification of lipid-binding proteins with unprecedented specificity. However, as the number of published lipid interactome studies grows, the lack of a centralized framework for comparing and integrating these datasets presents a significant challenge. The Lipid Interactome was developed to address this need by providing researchers with a structured platform to explore, cross-reference, and build upon previous findings. By aggregating and standardizing lipid–protein interaction data, this resource facilitates the rapid identification of candidate lipid-binding proteins and supports orthogonal validation efforts.

Designed as a continuously expanding repository, the Lipid Interactome will evolve alongside advances in lipid probe technologies and interactomics. Future iterations will incorporate validation studies for each lipid probe, offering researchers a dedicated space to compare findings across experimental contexts. This functionality will not only reduce redundancy in experimental design but also strengthen the reproducibility and biological interpretation of lipid interactome research. To support this vision, we invite researchers to contribute their datasets via our [Data Submission page](https://www.lipidinteractome.org/DataSubmission.html), fostering a collaborative and transparent approach to lipid biology.

## Abbreviations

UV
: *U*ltra *V*iolet

MS
: *M*ass *S*pectrometry

LIMMA
: *Li*near *M*odels for *M*icro*a*rray and RNA-Seq Data

ANOVA
: *An*alysis *o*f *Va*riance
